---
title: "Species Selection to Minimize Biomass Variability"
author: "Julian Olaya-Restrepo"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    keep_tex: false
    latex_engine: xelatex
geometry: margin=1in
fontsize: 11pt
documentclass: article
---
# BIOMASS

To reduce variability in aggregated fish biomass data for boosted regression tree modeling, we applied a multi-criteria selection framework using data from 2025 only (the year used for model training). Our objective was to identify the most stable and representative species within each taxonomic family while maintaining ecological realism and ensuring all families were adequately represented.


# Exploratory Visualization ----------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 7,
  fig.height = 5,
  dpi = 300,
  out.width = '85%'
)

# Load all required packages
library(dplyr)
library(tidyr)
library(purrr)
library(rvc)
library(ggplot2)
library(ggrepel)
library(lme4)
library(knitr)
library(xtable)
library(corrplot)

# Set options
options(scipen = 999)
```


# Data Preparation

## Download Data 

```{r, message=FALSE, warning=FALSE, results='hide'}
# Get all USVI data for specified years and regions
USVI <- getRvcData(years = c(2017:2025), regions = c("STTSTJ", "STX")) 
```

## Construct species lookup table 
 Obtain taxonomic information from all species using the rcv package.
 
 https://jeremiaheb.github.io/NCRMP_R_Fish_Package/
 
```{r pressure}
# 3. Group (Family) Table Construction ----------------------------------------
species_group_list <- "
Myripristis jacobus,Holocentridae
Apogon aurolineatus,Apogonidae
Apogon binotatus,Apogonidae
Apogon lachneri,Apogonidae
Apogon maculatus,Apogonidae
Apogon phenax,Apogonidae
Apogon pseudomaculatus,Apogonidae
Apogon quadrisquamatus,Apogonidae
Apogon townsendi,Apogonidae
Haemulon flavolineatum,Haemulidae
Sparisoma aurofrenatum,Scaridae
Lutjanus apodus,Lutjanidae
Lutjanus jocu,Lutjanidae
Lutjanus mahogoni,Lutjanidae
Epinephelus guttatus,Epinephelidae
Chaetodon capistratus,Chaetodontidae
Chaetodon striatus,Chaetodontidae
Chaetodon ocellatus,Chaetodontidae
Chaetodon sedentarius,Chaetodontidae
Sparisoma viride,Scaridae
Microspathodon chrysurus,Pomacentridae
Canthigaster rostrata,Tetraodontidae
Stegastes planifrons,Pomacentridae
Stegastes partitus,Pomacentridae
Stegastes adustus,Pomacentridae
Stegastes leucostictus,Pomacentridae
Chromis cyanea,Pomacentridae
Chromis multilineata,Pomacentridae
Coryphopterus personatus,Gobiidae
Coryphopterus glaucofraenum,Gobiidae
Gobiosoma grosvenori,Gobiidae
Holocentrus rufus,Holocentridae
Thalassoma bifasciatum,Labridae
Halichoeres bivittatus,Labridae
Halichoeres burekae,Labridae
Halichoeres caudalis,Labridae
Halichoeres cyanocephalus,Labridae
Halichoeres garnoti,Labridae
Halichoeres maculipinna,Labridae
Halichoeres pictus,Labridae
Halichoeres poeyi,Labridae
Halichoeres radiatus,Labridae
Scarus iseri,Scaridae
"
group_df <- read.csv(text = species_group_list, header = FALSE, col.names = c("SPECIES", "group"), strip.white=TRUE)

# 4. Match SPECIES to Internal Code -------------------------------------------
# Get RVC taxonomic codes (SPECIES_CD) for all analysis species
sp_table <- USVI$taxonomic_data
group_df2 <- group_df %>% 
  left_join(sp_table %>% select(SPECIES_CD, SCINAME), by = c("SPECIES" = "SCINAME")) %>%
  filter(!is.na(SPECIES_CD)) # keep only matched species

head(sp_table)
```

## Calculate Biomass Estimates


```{r, message=FALSE, warning=FALSE}
# Calculate mean biomass per PSU per species (no group parameter, for species-level)
gpsu_bio_species <- getPSUBiomass(
  USVI,
  species = group_df2$SPECIES_CD
)

# Add group and environmental data
gpsu_bio_species <- gpsu_bio_species %>%
  left_join(group_df2 %>% select(SPECIES_CD, group, SPECIES), by = "SPECIES_CD")

psu_env <- USVI$sample_data %>%
  select(PRIMARY_SAMPLE_UNIT, LAT_DEGREES, LON_DEGREES, DEPTH) %>%
  distinct(PRIMARY_SAMPLE_UNIT, .keep_all = TRUE)

final_gpsu <- gpsu_bio_species %>%
  left_join(psu_env, by = "PRIMARY_SAMPLE_UNIT") %>%
  select(YEAR, REGION, STRAT, PROT, PRIMARY_SAMPLE_UNIT,
         SPECIES, SPECIES_CD, group, m, var, biomass,
         LAT_DEGREES, LON_DEGREES, DEPTH)

head(final_gpsu)
```


# Exploratory Visualization

## Temporal Trends by Region

```{r, message=FALSE, warning=FALSE}
#Summarize for Visualization (Species-Level within Groups) ----------
plot_bio <- final_gpsu %>%
  group_by(REGION, YEAR, group, SPECIES_CD) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Loop by region and plot one at a time
for (reg in unique(plot_bio$REGION)) {
  plot_bio_reg <- plot_bio %>% filter(REGION == reg)
  
  label_data_reg <- plot_bio_reg %>%
    group_by(group, SPECIES_CD) %>%
    filter(YEAR == max(YEAR)) %>%
    ungroup()
  
  p <- ggplot(plot_bio_reg, aes(x = YEAR, y = mean_biomass, color = SPECIES_CD, group = SPECIES_CD)) +
    geom_line() +
    geom_point(size = 2) +
    geom_text_repel(data = label_data_reg, aes(label = SPECIES_CD),
                    nudge_x = 0.2, hjust = 0, size = 2.5, max.overlaps = 20) +
    scale_x_continuous(breaks = c(2017, 2019, 2021, 2023, 2025)) +
    facet_wrap(~ group, scales = "free_y") +
    theme_classic() +
    labs(y = "Biomass (kg per cylinder)", x = "Year",
         title = paste("Temporal Trends -", reg)) +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)
}
```

# 2025 Family-Level Biomass

## General plot by group

 
```{r, message=FALSE, warning=FALSE}
# 1. Group by REGION and group for 2025
bar_group <- final_gpsu %>%
  filter(YEAR == 2025) %>%
  group_by(REGION, group) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    se_biomass = sd(biomass, na.rm = TRUE) / sqrt(sum(!is.na(biomass))),
    .groups = "drop"
  ) %>%
  group_by(group) %>%
  mutate(overall_mean = mean(mean_biomass)) %>%
  ungroup() %>%
  mutate(group = reorder(group, -overall_mean))

ggplot(bar_group, aes(x = group, y = mean_biomass, fill = REGION)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_biomass - se_biomass, ymax = mean_biomass + se_biomass),
                position = position_dodge(0.8), width = 0.25) +
  scale_fill_manual(values = c("grey60", "steelblue")) +
  theme_classic() +
  labs(y = "Mean Biomass (kg per cylinder)", x = "Family",
       title = "Family Biomass in 2025 by Region") +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
```

```{r, message=FALSE, warning=FALSE}
# Aggregate: one value per REGION, group, and PSU (mean biomass of group per PSU for that region)
box_data <- final_gpsu %>%
  filter(YEAR == 2025) %>%
  group_by(REGION, group, PRIMARY_SAMPLE_UNIT) %>%
  summarise(group_biomass = mean(biomass, na.rm = TRUE), .groups = "drop")

# Order groups by median biomass across all regions, for left-to-right order
# (Same order for both regions, helps comparison)
median_group <- box_data %>%
  group_by(group) %>%
  summarise(med = median(group_biomass, na.rm = TRUE)) %>%
  arrange(desc(med))
box_data$group <- factor(box_data$group, levels = median_group$group)

ggplot(box_data, aes(x = group, y = group_biomass, fill = REGION)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(0.8), width = 0.6, alpha = 0.7) +
  geom_jitter(aes(color = REGION), size = 1, alpha = 0.5,
              position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.8)) +
  scale_y_log10() +
  theme_classic() +
  labs(y = "Biomass (kg, log10) per PSU", x = "Family",
       title = "Distribution of Family Biomass by Region (2025)") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r, message=FALSE, warning=FALSE, fig.height=18, fig.width=18}
# New name for the boxplot data to avoid confusion
box_data_species <- final_gpsu %>%
  filter(YEAR == 2025) %>%
  group_by(group) %>%
  mutate(SPECIES_CD = reorder(SPECIES_CD, -biomass, FUN = median, na.rm = TRUE)) %>%
  ungroup()

ggplot(box_data_species, aes(x = SPECIES_CD, y = biomass, fill = REGION)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(0.75), width = 0.6, alpha = 0.7) +
  geom_jitter(aes(color = REGION), size = 0.8, alpha = 0.5,
              position = position_jitterdodge(jitter.width = 0.10, dodge.width = 0.75)) +
  scale_y_log10() +
  facet_wrap(~ group, scales = "free_x") +
  theme_classic() +
  labs(y = "Biomass (kg, log10)", x = "Species",
       title = "Species Biomass Distribution by Region (2025)") +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        strip.text = element_text(size = 11),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

\newpage

# Model Dataset Creation

```{r}
# ============================================================================
# TOTAL REEF FISH BIOMASS AGGREGATION - ALL SPECIES (2025)
# ============================================================================

# Aggregate biomass for ALL species to get single value per PSU
total_biomass_2025 <- data_2025 %>%
  group_by(PRIMARY_SAMPLE_UNIT, REGION, STRAT, PROT, 
           LAT_DEGREES, LON_DEGREES, DEPTH) %>%
  summarise(
    # Total biomass in kg (sum of ALL species)
    total_biomass_kg = sum(biomass, na.rm = TRUE),
    
    # Number of species present (biomass > 0)
    n_species_present = n_distinct(SPECIES_CD[biomass > 0]),
    
    # Total number of species observed
    n_species_observed = n(),
    
    .groups = "drop"
  ) %>%
  mutate(
    # Unit conversions
    area_m2 = pi * (7.5^2),
    
    # Biomass in kg/m²
    biomass_kg_m2 = round(total_biomass_kg / area_m2, 4),
    
    # Biomass in g/m²
    biomass_g_m2 = round((total_biomass_kg / area_m2) * 1000, 2),
    
    # Biomass in t/km²
    biomass_t_km2 = round((total_biomass_kg / area_m2) * 1000, 2)
  ) %>%
  arrange(REGION, PRIMARY_SAMPLE_UNIT)

# Summary statistics
cat("=== TOTAL BIOMASS SUMMARY (ALL SPECIES) ===\n\n")
cat("Total stations:", nrow(total_biomass_2025), "\n")
cat("Regions:", paste(unique(total_biomass_2025$REGION), collapse = ", "), "\n\n")

# Overall summary
overall_stats <- total_biomass_2025 %>%
  summarise(
    mean_biomass = round(mean(total_biomass_kg, na.rm = TRUE), 2),
    sd_biomass = round(sd(total_biomass_kg, na.rm = TRUE), 2),
    median_biomass = round(median(total_biomass_kg, na.rm = TRUE), 2),
    min_biomass = round(min(total_biomass_kg, na.rm = TRUE), 2),
    max_biomass = round(max(total_biomass_kg, na.rm = TRUE), 2),
    mean_species = round(mean(n_species_present, na.rm = TRUE), 1)
  )

print(overall_stats)

# By region
cat("\n=== SUMMARY BY REGION ===\n")
region_stats <- total_biomass_2025 %>%
  group_by(REGION) %>%
  summarise(
    n_stations = n(),
    mean_biomass = round(mean(total_biomass_kg, na.rm = TRUE), 2),
    sd_biomass = round(sd(total_biomass_kg, na.rm = TRUE), 2),
    median_biomass = round(median(total_biomass_kg, na.rm = TRUE), 2),
    mean_species = round(mean(n_species_present, na.rm = TRUE), 1),
    .groups = "drop"
  )

print(region_stats)

# Export to CSV
write.csv(total_biomass_2025, 
          "total_reef_fish_biomass_2025.csv", 
          row.names = FALSE)

cat("\n=== DATASET EXPORTED ===\n")
cat("File: total_reef_fish_biomass_2025.csv\n")
cat("Rows (PSUs):", nrow(total_biomass_2025), "\n")
cat("Columns:", ncol(total_biomass_2025), "\n\n")

# Preview
cat("=== FIRST 10 ROWS ===\n")
total_biomass_2025 %>%
  select(PRIMARY_SAMPLE_UNIT, REGION, total_biomass_kg, 
         biomass_g_m2, n_species_present) %>%
  head(10) %>%
  print()
```



\newpage

# Statistical Analyses for Species variance

```{r}
# Filter to 2025 data only
data_2025 <- final_gpsu %>%
  filter(YEAR == 2025)

# Check sample size
cat("Total observations in 2025:", nrow(data_2025), "\n")
cat("Unique PSUs:", n_distinct(data_2025$PRIMARY_SAMPLE_UNIT), "\n")
cat("Species analyzed:", n_distinct(data_2025$SPECIES_CD), "\n")
```

## Coefficient of Variation (CV) Analysis

The coefficient of variation (CV = [standard deviation / mean] × 100) quantifies relative variability of biomass for each species across all sampling units. Lower CV values indicate more consistent biomass estimates, making species more suitable for predictive modeling. Species with CV >350% exhibit high spatial instability that can propagate error when biomass is aggregated to the family level.

**Interpretation**: A species with CV = 50% shows moderate variability, while CV > 200% indicates biomass varies more than twice the mean, suggesting the species is either rare, highly patchy, or experiences boom-bust dynamics.

```{r, message=FALSE, warning=FALSE}
# Calculate CV for each species
species_cv <- data_2025 %>%
  group_by(group, SPECIES_CD, SPECIES) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    sd_biomass = sd(biomass, na.rm = TRUE),
    cv = (sd_biomass / mean_biomass) * 100,
    n_obs = sum(!is.na(biomass)),
    .groups = "drop"
  )

ggplot(species_cv, aes(x = reorder(SPECIES_CD, cv), y = cv, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 350, linetype = "dashed", color = "red") +
  facet_wrap(~ group, scales = "free") +
  coord_flip() +
  theme_classic() +
  labs(y = "CV (%)", x = "Species", title = "Coefficient of Variation (2025)") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 2. Frequency of Occurrence 

Frequency of occurrence was calculated as the percentage of sampling units where a species was present (biomass > 0). Higher frequencies indicate ubiquitous species that consistently contribute to community biomass and are more reliably detected across the study area. Species present in <15% of samples are typically too rare to estimate biomass reliably or predict spatial patterns.

```{r, message=FALSE, warning=FALSE}
# Calculate frequency of occurrence
species_frequency <- data_2025 %>%
  group_by(group, SPECIES_CD, SPECIES) %>%
  summarise(
    n_obs = n(),
    n_present = sum(biomass > 0),
    freq_occurrence = (n_present / n_obs) * 100,
    .groups = "drop"
  )

ggplot(species_frequency, aes(x = reorder(SPECIES_CD, freq_occurrence), 
                              y = freq_occurrence, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red") +
  facet_wrap(~ group, scales = "free") +
  coord_flip() +
  theme_classic() +
  labs(y = "Frequency (%)", x = "Species", title = "Frequency of Occurrence (2025)") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold"))
```

## Contribution to Family Biomass

Each species' proportional contribution to family-level biomass was calculated as: (mean species biomass / mean family biomass) × 100 within the 2025 dataset. This identifies dominant species that drive family-level patterns. Species contributing >30% of family biomass were considered ecologically dominant, while those contributing <10% were considered minor contributors.

```{r, message=FALSE, warning=FALSE}
# Calculate contribution to family biomass
species_contribution <- data_2025 %>%
  group_by(group, SPECIES_CD, SPECIES) %>%
  summarise(
    species_mean_biomass = mean(biomass, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(group) %>%
  mutate(
    group_total = sum(species_mean_biomass, na.rm = TRUE),
    mean_contribution = (species_mean_biomass / group_total) * 100
  ) %>%
  ungroup() %>%
  arrange(group, desc(mean_contribution))

species_pareto <- species_contribution %>%
  group_by(group) %>%
  arrange(desc(mean_contribution)) %>%
  mutate(cumulative_contrib = cumsum(mean_contribution)) %>%
  ungroup()

ggplot(species_pareto, aes(x = reorder(SPECIES_CD, -mean_contribution), 
                           y = cumulative_contrib, color = group, group = group)) +
  geom_line(size = 1) +
  geom_point(aes(size = mean_contribution), alpha = 0.7) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "red") +
  facet_wrap(~ group, scales = "free", ncol = 3) +
  coord_flip() +
  theme_classic() +
  labs(y = "Cumulative Contribution (%)", x = "Species", 
       title = "Cumulative Biomass Contribution (2025)") +
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.y = element_text(size = 7))
```

## Correlation Analysis

Pairwise Pearson correlations of species biomass across sampling units identified redundant species with similar spatial patterns. Species pairs with correlation coefficients >0.7 were flagged as potentially redundant, as highly correlated species provide minimal independent information but increase model complexity.


```{r, message=FALSE, warning=FALSE}
biomass_wide <- data_2025 %>%
  select(PRIMARY_SAMPLE_UNIT, SPECIES_CD, biomass) %>%
  pivot_wider(names_from = SPECIES_CD, values_from = biomass, values_fill = 0)

cor_matrix <- cor(biomass_wide %>% select(-PRIMARY_SAMPLE_UNIT), use = "pairwise.complete.obs")
cor_matrix_clean <- cor_matrix
cor_matrix_clean[is.na(cor_matrix_clean)] <- 0

corrplot(cor_matrix_clean, 
         method = "circle", 
         type = "upper",
         order = "original",
         tl.cex = 0.5, 
         tl.col = "black",
         col = colorRampPalette(c("#6D9EC1", "white", "#E46726"))(200),
         title = "Species Correlation Matrix (2025)",
         mar = c(0,0,1,0),
         diag = FALSE)
```
Correlation analysis revealed no species pairs with correlation coefficients exceeding |r| = 0.7, indicating that species exhibit largely independent spatial patterns across the study area. A total of 231 pairwise comparisons (29.6%) resulted in undefined correlations due to species having zero variance (constant or absent across all sites), further confirming the presence of hyper-rare, uninformative species in the dataset. Only two species (*Stegastes partitus* and *Thalassoma bifasciatum*) showed moderate correlation (|r| > 0.5) with other species, both of which were retained in the final selection due to their high quality metrics. The lack of strong inter-species correlations confirmed that removing low-quality species based on variability criteria would not result in loss of independent spatial information.



## Zero-Inflation Analysis

The proportion of zero observations (absence records) was calculated for each species. High zero-inflation (>80%) indicates rare or highly patchy distributions that inflate variance and complicate statistical modeling. Zero-inflated species often require specialized models and may not aggregate well with continuously present species.

```{r, message=FALSE, warning=FALSE}
# Calculate proportion of zeros
zero_inflation <- data_2025 %>%
  group_by(group, SPECIES_CD, SPECIES) %>%
  summarise(
    n_obs = n(),
    n_zeros = sum(biomass == 0),
    prop_zeros = (n_zeros / n_obs) * 100,
    mean_when_present = mean(biomass[biomass > 0], na.rm = TRUE),
    .groups = "drop"
  )

ggplot(zero_inflation, aes(x = prop_zeros, y = mean_when_present, 
                           color = group, label = SPECIES_CD)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_vline(xintercept = 80, linetype = "dashed", color = "red") +
  geom_text_repel(size = 2, max.overlaps = 15) +
  scale_y_log10() +
  theme_classic() +
  labs(x = "% Zeros", y = "Mean Biomass When Present (kg, log10)",
       title = "Zero-Inflation Analysis (2025)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "bottom")
```

## Taylor's Power Law

We examined the mean-variance relationship following Taylor's Power Law: log(variance) = log(a) + b × log(mean). The slope parameter (b) indicates spatial aggregation patterns, where b ≈ 2 suggests random distribution, b > 2 indicates patchy/aggregated distribution, and b < 2 suggests regular spacing. Species deviating strongly from this relationship represent outliers in their distributional ecology.

```{r, message=FALSE, warning=FALSE}
# Combine mean and variance data
taylor_data <- species_cv %>%
  mutate(
    variance = sd_biomass^2,
    log_mean = log10(mean_biomass + 0.001),
    log_var = log10(variance + 0.001)
  ) %>%
  filter(is.finite(log_mean) & is.finite(log_var))

taylor_model <- lm(log_var ~ log_mean, data = taylor_data)
slope <- coef(taylor_model)[2]
r_squared <- summary(taylor_model)$r.squared

ggplot(taylor_data, aes(x = log_mean, y = log_var, color = group, label = SPECIES_CD)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  geom_text_repel(size = 2, max.overlaps = 10) +
  annotate("text", 
           x = min(taylor_data$log_mean, na.rm = TRUE) + 0.5, 
           y = max(taylor_data$log_var, na.rm = TRUE) - 0.5,
           label = paste0("slope = ", round(slope, 2), 
                         "\nR-squared = ", round(r_squared, 3)),
           hjust = 0, size = 4, fontface = "bold") +
  theme_classic() +
  labs(x = "log10(Mean Biomass)", y = "log10(Variance)",
       title = "Taylor's Power Law (2025)",
       subtitle = "Species above the line are more variable than expected") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 9),
        legend.position = "bottom")

```

\newpage

# Species Selection Strategy

## Integrate All Metrics

```{r}
species_metrics <- species_cv %>%
  left_join(species_frequency %>% select(SPECIES_CD, freq_occurrence), by = "SPECIES_CD") %>%
  left_join(species_contribution %>% select(SPECIES_CD, mean_contribution), by = "SPECIES_CD") %>%
  left_join(zero_inflation %>% select(SPECIES_CD, prop_zeros), by = "SPECIES_CD") %>%
  mutate(
    selection_score = 
      (100 - pmin(cv, 500)) * 0.30 +
      freq_occurrence * 0.25 +
      mean_contribution * 0.25 +
      (100 - prop_zeros) * 0.20
  )
```

## Three-Tier Selection Strategy

```{r}
# Tier 1: High quality
tier1 <- species_metrics %>%
  filter(!is.na(cv), cv < 350, freq_occurrence > 15, 
         mean_contribution > 10, prop_zeros < 80) %>%
  mutate(selection_method = "High quality")

# Tier 2: Dominant
tier2 <- species_metrics %>%
  filter(!is.na(cv)) %>%
  anti_join(tier1, by = "SPECIES_CD") %>%
  filter(mean_contribution > 30) %>%
  mutate(selection_method = "Dominant")

# Tier 3: Best available
tier3 <- species_metrics %>%
  filter(!is.na(cv)) %>%
  anti_join(bind_rows(tier1, tier2), by = "SPECIES_CD") %>%
  group_by(group) %>%
  slice_max(order_by = selection_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(selection_method = "Best available")

# Combine and remove hyper-rare Apogonids
final_selection <- bind_rows(tier1, tier2, tier3) %>%
  filter(!(SPECIES %in% c("Apogon binotatus", "Apogon quadrisquamatus")))
final_selection
```

Final Selection: 24 species across 11 families were selected:

- Tier 1 (High Quality, n=16): Mean CV = 201%, mean frequency = 62%
- Tier 2 (Dominant, n=1): Lutjanus apodus (74% of family biomass)
- Tier 3 (Best Available, n=7): Representatives for remaining families
Two hyper-rare Apogonids (CV >2000%, <2% occurrence) were excluded.


## Variance Reduction Validation

```{r, message=FALSE, warning=FALSE}
selected_species_codes <- final_selection$SPECIES_CD

all_species_biomass <- data_2025 %>%
  group_by(PRIMARY_SAMPLE_UNIT, group, REGION) %>%
  summarise(family_biomass = sum(biomass, na.rm = TRUE), .groups = "drop") %>%
  mutate(dataset = "All species (n=40)")

selected_species_biomass <- data_2025 %>%
  filter(SPECIES_CD %in% selected_species_codes) %>%
  group_by(PRIMARY_SAMPLE_UNIT, group, REGION) %>%
  summarise(family_biomass = sum(biomass, na.rm = TRUE), .groups = "drop") %>%
  mutate(dataset = "Selected species (n=24)")

comparison_data <- bind_rows(all_species_biomass, selected_species_biomass)

variance_comparison <- comparison_data %>%
  group_by(group, dataset) %>%
  summarise(
    mean_biomass = mean(family_biomass, na.rm = TRUE),
    sd_biomass = sd(family_biomass, na.rm = TRUE),
    cv = (sd_biomass / mean_biomass) * 100,
    .groups = "drop"
  )

variance_reduction <- variance_comparison %>%
  select(group, dataset, cv, sd_biomass) %>%
  pivot_wider(names_from = dataset, values_from = c(cv, sd_biomass)) %>%
  mutate(cv_reduction_pct = ((`cv_All species (n=40)` - `cv_Selected species (n=24)`) / 
                              `cv_All species (n=40)`) * 100)

biomass_retention <- comparison_data %>%
  select(PRIMARY_SAMPLE_UNIT, group, dataset, family_biomass) %>%
  pivot_wider(names_from = dataset, values_from = family_biomass) %>%
  group_by(group) %>%
  summarise(
    total_biomass_all = sum(`All species (n=40)`, na.rm = TRUE),
    total_biomass_selected = sum(`Selected species (n=24)`, na.rm = TRUE),
    biomass_retained_pct = (total_biomass_selected / total_biomass_all) * 100,
    .groups = "drop"
  )

ggplot(variance_comparison, aes(x = reorder(group, -cv), y = cv, fill = dataset)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme_classic() +
  labs(x = "Family", y = "CV (%)", title = "CV: All vs Selected Species",
       fill = "Dataset") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), legend.position = "bottom") +
  scale_fill_manual(values = c("All species (n=40)" = "#E69F00", 
                                "Selected species (n=24)" = "#56B4E9"))
```
```{r}
biomass_comparison_wide <- comparison_data %>%
  select(PRIMARY_SAMPLE_UNIT, group, dataset, family_biomass) %>%
  pivot_wider(names_from = dataset, values_from = family_biomass)

ggplot(biomass_comparison_wide, 
       aes(x = `All species (n=40)`, y = `Selected species (n=24)`)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  facet_wrap(~ group, scales = "free") +
  theme_classic() +
  labs(x = "Biomass - All Species (kg)", y = "Biomass - Selected Species (kg)",
       title = "Family Biomass: All vs Selected Species") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
Variance Reduction Results:

Mean CV reduction: 7.5% (317% → 294%)
Overall biomass retention: 96.2%
Families improved: Apogonidae (21% CV reduction)
Families with slight CV increases: 4 families (+0.4% to +7.8%), all retaining >86% biomass

\newpage

# Results Summary

## Species Selection and Variance Reduction -----------------------------------

Analysis of 40 candidate species using Taylor's Power Law (R² = 0.83, slope = 0.92) revealed regular spatial spacing patterns typical of territorial reef fishes. Species exhibited high natural variability (median CV = 322%) and zero-inflation (median 70% zeros), confirming inherent patchiness.

The three-tier selection strategy identified 24 species:

**Tier 1 (High Quality, n=16):** Including Sparisoma aurofrenatum (CV=111%), Thalassoma bifasciatum (CV=137%), Halichoeres garnoti (CV=143%), and Stegastes partitus (CV=195%). These species provide stable biomass estimates (mean CV=201%, mean frequency=62%, mean contribution=51%).

**Tier 2 (Dominant, n=1):** Lutjanus apodus contributes 74% of Lutjanidae biomass despite moderate variability (CV=394%).

**Tier 3 (Best Available, n=7):** Representatives ensuring complete family coverage, including Apogon maculatus, Chaetodon ocellatus, and Coryphopterus glaucofraenum.

Two hyper-rare Apogonids (*A. binotatus*, *A. quadrisquamatus*) with CV >2000% and <2% occurrence were excluded.

Species selection achieved 7.5% CV reduction while retaining 96.2% of total biomass. The modest variance reduction reflects intrinsic reef fish patchiness rather than selection failure. No species pairs showed r >0.7, confirming independent spatial information. The final selection balances ecological representativeness, statistical stability, and model parsimony.

## Summary Table

```{r}
library(knitr)

pub_table <- variance_reduction %>%
  left_join(biomass_retention %>% select(group, biomass_retained_pct), by = "group") %>%
  left_join(species_metrics %>% count(group, name = "n_total"), by = "group") %>%
  left_join(final_selection %>% count(group, name = "n_selected"), by = "group") %>%
  transmute(
    Family = group,
    `Species (all/sel)` = paste0(n_total, "/", n_selected),
    `CV All` = round(`cv_All species (n=40)`, 1),
    `CV Selected` = round(`cv_Selected species (n=24)`, 1),
    `CV Reduction` = round(cv_reduction_pct, 1),
    `Biomass Retained` = round(biomass_retained_pct, 1)
  )

# Add overall row
overall <- data.frame(
  Family = "**Overall**",
  `Species (all/sel)` = "**40/24**",
  `CV All` = round(mean(pub_table$`CV All`, na.rm = TRUE), 1),
  `CV Selected` = round(mean(pub_table$`CV Selected`, na.rm = TRUE), 1),
  `CV Reduction` = round((mean(pub_table$`CV All`, na.rm = TRUE) - 
                   mean(pub_table$`CV Selected`, na.rm = TRUE)) / 
                  mean(pub_table$`CV All`, na.rm = TRUE) * 100, 1),
  `Biomass Retained` = round(sum(biomass_retention$total_biomass_selected, na.rm = TRUE) / 
                   sum(biomass_retention$total_biomass_all, na.rm = TRUE) * 100, 1),
  check.names = FALSE
)

pub_table_final <- rbind(pub_table, overall)

# Print with kable - THIS WILL WORK
kable(pub_table_final,
      booktabs = TRUE,
      caption = "Summary of species selection criteria and variance reduction by family. CV = Coefficient of Variation (%). Negative CV reduction indicates slight increase. Families with CV increases retained >86% of biomass.",
      align = c("l", "c", "c", "c", "c", "c"))
```

# Export dataset
```{r}
# Get the 24 selected species codes
selected_species_codes <- final_selection$SPECIES_CD

# Filter data_2025 to only selected species
final_dataset <- data_2025 %>%
  filter(SPECIES_CD %in% selected_species_codes) %>%
  # Add selection tier information
  left_join(final_selection %>% select(SPECIES_CD, selection_method, 
                                       cv, freq_occurrence, mean_contribution, 
                                       prop_zeros, selection_score),
            by = "SPECIES_CD") %>%
  # Arrange for clarity
  arrange(group, SPECIES_CD, PRIMARY_SAMPLE_UNIT)

# Export to CSV
write.csv(final_dataset, 
          "final_selected_species_data_2025.csv", 
          row.names = FALSE)

# Summary of exported data
cat("\n=== FINAL DATASET EXPORTED ===\n")
cat("File: final_selected_species_data_2025.csv\n")
cat("Total observations:", nrow(final_dataset), "\n")
cat("Unique species:", n_distinct(final_dataset$SPECIES_CD), "\n")
cat("Unique PSUs:", n_distinct(final_dataset$PRIMARY_SAMPLE_UNIT), "\n")
cat("Families represented:", n_distinct(final_dataset$group), "\n\n")

# Summary by species
cat("Records per species:\n")
final_dataset %>%
  count(group, SPECIES, selection_method) %>%
  arrange(group, SPECIES) %>%
  print(n = Inf)

# Quick preview
cat("\n=== DATASET PREVIEW ===\n")
head(final_dataset, 10)
```


```{r}
# ============================================================================
# GENERAL REEF FISH BIOMASS INDICATOR BY STATION (PSU) - 2025 (CORRECTED)
# ============================================================================

# 1. Verify current biomass units
cat("=== UNIT VERIFICATION ===\n")
cat("Current biomass unit in RVC data:\n")
cat("- getPSUBiomass() returns biomass in kg per sampled cylinder\n")
cat("- RVC standard cylinder area: π × (7.5m)² = 176.71 m²\n\n")

# 2. Filter selected species and calculate total biomass per PSU
biomass_indicator_2025 <- data_2025 %>%
  filter(SPECIES_CD %in% selected_species_codes) %>%
  group_by(PRIMARY_SAMPLE_UNIT, REGION, STRAT, PROT, 
           LAT_DEGREES, LON_DEGREES, DEPTH) %>%
  summarise(
    # Total biomass in kg (sum of all selected species)
    total_biomass_kg = sum(biomass, na.rm = TRUE),
    
    # Number of species present (biomass > 0)
    n_species_present = n_distinct(SPECIES_CD[biomass > 0]),
    
    # Total number of selected species observed at the PSU
    n_species_observed = n(),
    
    .groups = "drop"
  ) %>%
  mutate(
    # Unit conversions
    # RVC uses 15m diameter cylinders = 7.5m radius
    # Area = π × r² = π × 7.5² = 176.71 m²
    area_m2 = pi * (7.5^2),
    
    # Biomass in g/m²
    biomass_g_m2 = (total_biomass_kg * 1000) / area_m2,
    
    # Biomass in kg/m²
    biomass_kg_m2 = total_biomass_kg / area_m2,
    
    # Biomass in t/km² - CORRECTED
    # Step 1: kg/m² 
    # Step 2: Convert to tonnes: kg/m² ÷ 1000 = t/m²
    # Step 3: Scale to km²: t/m² × 1,000,000 = t/km²
    # Combined: (kg/m²) × 1000 = t/km²
    biomass_t_km2 = biomass_kg_m2 * 1000
  ) %>%
  arrange(REGION, PRIMARY_SAMPLE_UNIT)

# 3. Statistical summary by region
cat("=== BIOMASS STATISTICAL SUMMARY BY REGION ===\n\n")
summary_by_region <- biomass_indicator_2025 %>%
  group_by(REGION) %>%
  summarise(
    n_stations = n(),
    mean_biomass_kg = round(mean(total_biomass_kg, na.rm = TRUE), 3),
    sd_biomass_kg = round(sd(total_biomass_kg, na.rm = TRUE), 3),
    median_biomass_kg = round(median(total_biomass_kg, na.rm = TRUE), 3),
    mean_biomass_g_m2 = round(mean(biomass_g_m2, na.rm = TRUE), 2),
    mean_biomass_kg_m2 = round(mean(biomass_kg_m2, na.rm = TRUE), 4),
    mean_biomass_t_km2 = round(mean(biomass_t_km2, na.rm = TRUE), 1),
    min_biomass_kg = round(min(total_biomass_kg, na.rm = TRUE), 3),
    max_biomass_kg = round(max(total_biomass_kg, na.rm = TRUE), 3),
    .groups = "drop"
  )

print(summary_by_region)

# 4. Overall summary (all stations)
cat("\n=== OVERALL SUMMARY (ALL STATIONS) ===\n")
overall_summary <- biomass_indicator_2025 %>%
  summarise(
    total_stations = n(),
    mean_biomass_kg = round(mean(total_biomass_kg, na.rm = TRUE), 3),
    sd_biomass_kg = round(sd(total_biomass_kg, na.rm = TRUE), 3),
    median_biomass_kg = round(median(total_biomass_kg, na.rm = TRUE), 3),
    mean_biomass_g_m2 = round(mean(biomass_g_m2, na.rm = TRUE), 2),
    mean_biomass_kg_m2 = round(mean(biomass_kg_m2, na.rm = TRUE), 4),
    mean_biomass_t_km2 = round(mean(biomass_t_km2, na.rm = TRUE), 1),
    mean_species_richness = round(mean(n_species_present, na.rm = TRUE), 1)
  )

print(overall_summary)

# Verification of conversions
cat("\n=== CONVERSION VERIFICATION ===\n")
cat("Example with first station:\n")
example <- biomass_indicator_2025 %>% slice(1)
cat(sprintf("  Total biomass: %.4f kg\n", example$total_biomass_kg))
cat(sprintf("  Area: %.2f m²\n", example$area_m2))
cat(sprintf("  Biomass: %.4f kg/m²\n", example$biomass_kg_m2))
cat(sprintf("  Biomass: %.2f g/m²\n", example$biomass_g_m2))
cat(sprintf("  Biomass: %.2f t/km²\n\n", example$biomass_t_km2))
cat("Check: biomass_g_m2 should = biomass_kg_m2 × 1000\n")
cat(sprintf("  %.2f = %.4f × 1000? %s\n", 
            example$biomass_g_m2, 
            example$biomass_kg_m2,
            ifelse(abs(example$biomass_g_m2 - example$biomass_kg_m2*1000) < 0.01, "✓ YES", "✗ NO")))
cat("Check: biomass_t_km2 should = biomass_kg_m2 × 1000\n")
cat(sprintf("  %.2f = %.4f × 1000? %s\n", 
            example$biomass_t_km2, 
            example$biomass_kg_m2,
            ifelse(abs(example$biomass_t_km2 - example$biomass_kg_m2*1000) < 0.01, "✓ YES", "✗ NO")))

# 5. Export complete dataset
write.csv(biomass_indicator_2025, 
          "reef_fish_biomass_indicator_2025.csv", 
          row.names = FALSE)

cat("\n=== DATASET EXPORTED ===\n")
cat("File: reef_fish_biomass_indicator_2025.csv\n")
cat("Stations (PSUs):", nrow(biomass_indicator_2025), "\n")

# 6. Dataset preview
cat("\n=== FIRST 10 STATIONS (ALL UNITS) ===\n")
biomass_indicator_2025 %>%
  select(PRIMARY_SAMPLE_UNIT, REGION, total_biomass_kg, 
         biomass_kg_m2, biomass_g_m2, biomass_t_km2, n_species_present) %>%
  head(10) %>%
  print()
```

