---
title: "Species Selection to Minimize Biomass Variability"
author: "Julian Olaya-Restrepo"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    keep_tex: false
    latex_engine: pdflatex
    citation_package: natbib
  geometry: margin=1in
  fontsize: 11pt
  documentclass: article
bibliography: references.bib
csl: ecology.csl
---

To reduce variability in aggregated fish biomass data for boosted regression tree modeling, we applied a multi-criteria selection framework using data from 2025 only (the year used for model training). Our objective was to identify the most stable and representative species within each taxonomic family while maintaining ecological realism and ensuring all families were adequately represented.


# Exploratory Visualization ---------------------------------------------------

```{r setup, include=FALSE}
# Load all required packages
library(dplyr)     # for data manipulation
library(tidyr)     # for nest/unnest
library(purrr)     # for map functions
library(rvc)       # for RVC fish data tools
library(ggplot2)   # for plotting
library(ggrepel)   # for non-overlapping plot labels
library(lme4)      # for mixed models
```

## Download Data 

```{r, message=FALSE, warning=FALSE, results='hide'}
# Get all USVI data for specified years and regions
USVI <- getRvcData(years = c(2017:2025), regions = c("STTSTJ", "STX")) 
```

## Construct species lookup table 
 Obtain taxonomic information from all species using the rcv package.
 
 https://jeremiaheb.github.io/NCRMP_R_Fish_Package/
 
```{r pressure, echo=FALSE}
# 3. Group (Family) Table Construction ----------------------------------------
species_group_list <- "Myripristis jacobus,Holocentridae
Apogon aurolineatus,Apogonidae
Apogon binotatus,Apogonidae
Apogon lachneri,Apogonidae
Apogon maculatus,Apogonidae
Apogon phenax,Apogonidae
Apogon pseudomaculatus,Apogonidae
Apogon quadrisquamatus,Apogonidae
Apogon townsendi,Apogonidae
Haemulon flavolineatum,Haemulidae
Sparisoma aurofrenatum,Scaridae
Lutjanus apodus,Lutjanidae
Lutjanus jocu,Lutjanidae
Lutjanus mahogoni,Lutjanidae
Epinephelus guttatus,Epinephelidae
Chaetodon capistratus,Chaetodontidae
Chaetodon striatus,Chaetodontidae
Chaetodon ocellatus,Chaetodontidae
Chaetodon sedentarius,Chaetodontidae
Sparisoma viride,Scaridae
Microspathodon chrysurus,Pomacentridae
Canthigaster rostrata,Tetraodontidae
Stegastes planifrons,Pomacentridae
Stegastes partitus,Pomacentridae
Stegastes adustus,Pomacentridae
Stegastes leucostictus,Pomacentridae
Chromis cyanea,Pomacentridae
Chromis multilineata,Pomacentridae
Coryphopterus personatus,Gobiidae
Coryphopterus glaucofraenum,Gobiidae
Gobiosoma grosvenori,Gobiidae
Holocentrus rufus,Holocentridae
Thalassoma bifasciatum,Labridae
Halichoeres bivittatus,Labridae
Halichoeres burekae,Labridae
Halichoeres caudalis,Labridae
Halichoeres cyanocephalus,Labridae
Halichoeres garnoti,Labridae
Halichoeres maculipinna,Labridae
Halichoeres pictus,Labridae
Halichoeres poeyi,Labridae
Halichoeres radiatus,Labridae
Scarus iseri,Scaridae
"
group_df <- read.csv(text = species_group_list, header = FALSE, col.names = c("SPECIES", "group"), strip.white=TRUE)

# 4. Match SPECIES to Internal Code -------------------------------------------
# Get RVC taxonomic codes (SPECIES_CD) for all analysis species
sp_table <- USVI$taxonomic_data
group_df2 <- group_df %>% 
  left_join(sp_table %>% select(SPECIES_CD, SCINAME), by = c("SPECIES" = "SCINAME")) %>%
  filter(!is.na(SPECIES_CD)) # keep only matched species

head(sp_table)
```

## Filter information for our selected group of species 


```{r, message=FALSE, warning=FALSE}
# 5. Estimate Biomass at PSU Level for All Species ----------------------------
# Calculate mean biomass per PSU per species (no group parameter, for species-level)
gpsu_bio_species <- getPSUBiomass(
  USVI,
  species = group_df2$SPECIES_CD
)

# 6. Add Group & Environmental Data -------------------------------------------
# Attach group and species names, for reference and aggregation
gpsu_bio_species <- gpsu_bio_species %>%
  left_join(group_df2 %>% select(SPECIES_CD, group, SPECIES), by = "SPECIES_CD")

# Create a spatial/environmental lookup table (unique per PSU)
psu_env <- USVI$sample_data %>%
  select(PRIMARY_SAMPLE_UNIT, LAT_DEGREES, LON_DEGREES, DEPTH) %>%
  distinct(PRIMARY_SAMPLE_UNIT, .keep_all = TRUE) # ensures one row per PSU

# Join spatial/environmental info
final_gpsu <- gpsu_bio_species %>%
  left_join(psu_env, by = "PRIMARY_SAMPLE_UNIT")

# 7. Keep and Order Only Needed Columns ---------------------------------------
final_gpsu <- final_gpsu %>%
  select(
    YEAR, REGION, STRAT, PROT, PRIMARY_SAMPLE_UNIT,
    SPECIES, SPECIES_CD, group, m, var, biomass,
    LAT_DEGREES, LON_DEGREES, DEPTH
  )

head(final_gpsu)
```


# Plots -----------------------------------------------------------------------

```{r, message=FALSE, warning=FALSE}
#Summarize for Visualization (Species-Level within Groups) ----------
plot_bio <- final_gpsu %>%
  group_by(REGION, YEAR, group, SPECIES_CD) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Loop by region and plot one at a time
for (reg in unique(plot_bio$REGION)) {
  plot_bio_reg <- plot_bio %>% filter(REGION == reg)
  
  # Prepare label data (last year for each group/species)
  label_data_reg <- plot_bio_reg %>%
    group_by(group, SPECIES_CD) %>%
    filter(YEAR == max(YEAR)) %>%
    ungroup()
  
  p <- ggplot(plot_bio_reg, aes(x = YEAR, y = mean_biomass, color = SPECIES_CD, group = SPECIES_CD)) +
  geom_line() +
  geom_point(size = 2) +
  ggrepel::geom_text_repel(
    data = label_data_reg,
    aes(label = SPECIES_CD, color = SPECIES_CD),
    nudge_x = 0.2,
    direction = "y",
    hjust = 0,
    show.legend = FALSE,
    size = 3
  ) +
  scale_x_continuous(breaks = c(2017, 2019, 2021, 2023, 2025)) +
  facet_wrap(~ group, scales = "free_y") +
  theme_classic() +
  ylab("Biomass (kg per cylinder)") +
  xlab("Year") +
  ggtitle(paste("Temporal Trends in Biomass by Group and Species -", reg)) +
  theme(
    plot.title    = element_text(hjust = 0.5),
    legend.position = "none",
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 12) # add this line!
  )
  
  print(p)
  # Optionally, save:
  # ggsave(paste0("Temporal_Trends_", reg, ".png"), plot = p, width = 10, height = 7)
}
```

# Year 2025 -------------------------------------------------------------------

## General plot by group

 
```{r, message=FALSE, warning=FALSE}
# 1. Group by REGION and group for 2025
bar_group <- final_gpsu %>%
  filter(YEAR == 2025) %>%
  group_by(REGION, group) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    se_biomass   = sd(biomass, na.rm = TRUE) / sqrt(sum(!is.na(biomass))),
    .groups      = "drop"
  )

# For sorting and clear side-by-side plot, arrange group factor by overall mean across regions
bar_group <- bar_group %>%
  group_by(group) %>%
  mutate(overall_mean = mean(mean_biomass, na.rm=TRUE)) %>%
  ungroup() %>%
  mutate(group = reorder(group, -overall_mean))

p_bar_biomass <- ggplot(bar_group, aes(x = group, y = mean_biomass, fill = REGION)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_biomass - se_biomass, ymax = mean_biomass + se_biomass, group = REGION),
    position = position_dodge(width = 0.8),
    width = 0.25, size = 0.8
  ) +
  geom_text(
    aes(label = round(mean_biomass, 2), group = REGION),
    position = position_dodge(width = 0.8),
    vjust = -0.8, size = 3
  ) +
  theme_classic() +
  ylab("Mean Biomass (kg per cylinder)") +
  xlab("Family") +
  ggtitle("Group Biomass in 2025 by Region") +
  scale_fill_manual(values = c("grey60", "steelblue")) + # Or pick 2 colors you like
  theme(
    plot.title   = element_text(hjust = 0.5, size = 16),
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y  = element_text(size = 14),
    axis.title.x = element_text(size = 12, margin = margin(t = 10)),
    axis.title.y = element_text(size = 12, margin = margin(r = 10)),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 12)
  )
p_bar_biomass
```

```{r, message=FALSE, warning=FALSE}
# Aggregate: one value per REGION, group, and PSU (mean biomass of group per PSU for that region)
box_data <- final_gpsu %>%
  filter(YEAR == 2025) %>%
  group_by(REGION, group, PRIMARY_SAMPLE_UNIT) %>%
  summarise(
    group_biomass = mean(biomass, na.rm = TRUE),
    .groups = "drop"
  )

# Order groups by median biomass across all regions, for left-to-right order
# (Same order for both regions, helps comparison)
median_group <- box_data %>%
  group_by(group) %>%
  summarise(med = median(group_biomass, na.rm = TRUE)) %>%
  arrange(desc(med))
box_data$group <- factor(box_data$group, levels = median_group$group)

p_box_biomass <- ggplot(box_data, aes(x = group, y = group_biomass, fill = REGION)) +
  geom_boxplot(
    outlier.shape = NA, 
    position = position_dodge(width = 0.8),
    width = 0.6,
    alpha = 0.7
  ) +
  geom_jitter(
    aes(color = REGION),
    size = 1.2,
    alpha = 0.5,
    position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.6)
  ) +
  scale_y_log10() +
  ylab("Biomass (kg per cylinder, log10) per PSU") +
  xlab("Family") +
  ggtitle("Distribution of Group Biomass per PSU by Region (2025, log scale)") +
  theme_classic() +
  theme(
    plot.title   = element_text(hjust = 0.5, size = 12),
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y  = element_text(size = 12),
    axis.title.x = element_text(size = 12, margin = margin(t = 10)),
    axis.title.y = element_text(size = 12, margin = margin(r = 10)),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 12)
  )
p_box_biomass
```


```{r, message=FALSE, warning=FALSE, fig.height=18, fig.width=18}
# New name for the boxplot data to avoid confusion
box_data_species <- final_gpsu %>%
  filter(YEAR == 2025) %>%
  group_by(group) %>%
  mutate(SPECIES_CD = reorder(SPECIES_CD, -biomass, FUN = median, na.rm = TRUE)) %>%
  ungroup()

p_box_biomass_sp <- ggplot(box_data_species, aes(x = SPECIES_CD, y = biomass, fill = REGION)) +
  geom_boxplot(
    outlier.shape = NA,
    position = position_dodge(width = 0.75),
    width = 0.6,
    alpha = 0.7
  ) +
  geom_jitter(
    aes(color = REGION),
    size = 1.2,
    alpha = 0.5,
    position = position_jitterdodge(jitter.width = 0.10, dodge.width = 0.6)
  ) +
  scale_y_log10() +
  ylab("Biomass (kg per cylinder, log10)") +
  xlab("Species") +
  ggtitle("Distribution of Biomass by Species and Region Within Groups (2025)") +
  facet_wrap(~ group, scales = "free_x") +
  theme_classic() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 18),
    strip.text    = element_text(size = 14),
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 11),
    axis.title.x  = element_text(size = 16, margin = margin(t = 10)),
    axis.title.y  = element_text(size = 16, margin = margin(r = 10)),
    axis.text.y   = element_text(size = 14),
    legend.title  = element_text(size = 13),
    legend.text   = element_text(size = 12)
  )
print(p_box_biomass_sp)
```


# Statistical Analyses for Variance Reduction ---------------------------------

```{r}
# Filter to 2025 data only
data_2025 <- final_gpsu %>%
  filter(YEAR == 2025)

# Check sample size
cat("Total observations in 2025:", nrow(data_2025), "\n")
cat("Unique PSUs:", n_distinct(data_2025$PRIMARY_SAMPLE_UNIT), "\n")
cat("Species analyzed:", n_distinct(data_2025$SPECIES_CD), "\n")
```

## 1. Coefficient of Variation (CV) Analysis

The coefficient of variation (CV = [standard deviation / mean] × 100) quantifies relative variability of biomass for each species across all sampling units. Lower CV values indicate more consistent biomass estimates, making species more suitable for predictive modeling. Species with CV >350% exhibit high spatial instability that can propagate error when biomass is aggregated to the family level.

**Interpretation**: A species with CV = 50% shows moderate variability, while CV > 200% indicates biomass varies more than twice the mean, suggesting the species is either rare, highly patchy, or experiences boom-bust dynamics.

```{r, message=FALSE, warning=FALSE}
# Calculate CV for each species
species_cv <- data_2025 %>%
  group_by(group, SPECIES_CD, SPECIES) %>%
  summarise(
    mean_biomass = mean(biomass, na.rm = TRUE),
    sd_biomass = sd(biomass, na.rm = TRUE),
    cv = (sd_biomass / mean_biomass) * 100,
    n_obs = sum(!is.na(biomass)),
    .groups = "drop"
  ) %>%
  arrange(group, cv)

# Summary statistics
cat("\n=== COEFFICIENT OF VARIATION SUMMARY ===\n")
summary(species_cv$cv)

# Visualize CV by species
ggplot(species_cv, aes(x = reorder(SPECIES_CD, cv), y = cv, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 350, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 5, y = 400, label = "CV threshold = 350%", color = "red") +
  facet_wrap(~ group, scales = "free") +
  coord_flip() +
  theme_classic() +
  labs(title = "Coefficient of Variation by Species (2025)",
       subtitle = "Lower values indicate more stable biomass estimates",
       x = "Species", y = "CV (%)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10))
```

## 2. Frequency of Occurrence 

Frequency of occurrence was calculated as the percentage of sampling units where a species was present (biomass > 0). Higher frequencies indicate ubiquitous species that consistently contribute to community biomass and are more reliably detected across the study area. Species present in <15% of samples are typically too rare to estimate biomass reliably or predict spatial patterns.

```{r, message=FALSE, warning=FALSE}
# Calculate frequency of occurrence
species_frequency <- data_2025 %>%
  group_by(group, SPECIES_CD, SPECIES) %>%
  summarise(
    n_obs = n(),
    n_present = sum(biomass > 0),
    freq_occurrence = (n_present / n_obs) * 100,
    .groups = "drop"
  ) %>%
  arrange(group, desc(freq_occurrence))

# Summary statistics
cat("\n=== FREQUENCY OF OCCURRENCE SUMMARY ===\n")
summary(species_frequency$freq_occurrence)

# Visualize frequency
ggplot(species_frequency, aes(x = reorder(SPECIES_CD, freq_occurrence), 
                              y = freq_occurrence, fill = group)) +
  geom_col() +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 5, y = 20, label = "Frequency threshold = 15%", color = "red") +
  facet_wrap(~ group, scales = "free") +
  coord_flip() +
  theme_classic() +
  labs(title = "Frequency of Occurrence by Species (2025)",
       subtitle = "Percentage of sampling units where species was present",
       x = "Species", y = "Frequency (%)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10))
```

## 3. Contribution to Total Biomass

Each species' proportional contribution to family-level biomass was calculated as: (mean species biomass / mean family biomass) × 100 within the 2025 dataset. This identifies dominant species that drive family-level patterns. Species contributing >30% of family biomass were considered ecologically dominant, while those contributing <10% were considered minor contributors.

```{r, message=FALSE, warning=FALSE}
# Calculate contribution to family biomass
species_contribution <- data_2025 %>%
  group_by(group, SPECIES_CD, SPECIES) %>%
  summarise(
    species_mean_biomass = mean(biomass, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(group) %>%
  mutate(
    group_total = sum(species_mean_biomass, na.rm = TRUE),
    mean_contribution = (species_mean_biomass / group_total) * 100
  ) %>%
  ungroup() %>%
  arrange(group, desc(mean_contribution))

# Verify contributions sum to 100% per group
cat("\n=== CONTRIBUTION VERIFICATION (should be ~100% per group) ===\n")
species_contribution %>%
  group_by(group) %>%
  summarise(total_contrib = sum(mean_contribution)) %>%
  print()

# Summary statistics
cat("\n=== CONTRIBUTION SUMMARY ===\n")
summary(species_contribution$mean_contribution)

# Pareto plot: cumulative contribution
species_pareto <- species_contribution %>%
  group_by(group) %>%
  arrange(desc(mean_contribution)) %>%
  mutate(cumulative_contrib = cumsum(mean_contribution)) %>%
  ungroup()

ggplot(species_pareto, aes(x = reorder(SPECIES_CD, -mean_contribution), 
                           y = cumulative_contrib, color = group, group = group)) +
  geom_line(size = 1.2) +
  geom_point(aes(size = mean_contribution), alpha = 0.7) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 5, y = 85, label = "80% of biomass", color = "red") +
  facet_wrap(~ group, scales = "free_x") +
  coord_flip() +
  theme_classic() +
  labs(title = "Cumulative Biomass Contribution (2025)",
       subtitle = "80/20 rule: few species account for most biomass",
       x = "Species", y = "Cumulative Contribution (%)") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10))
```

## 4. Correlation Analysis

Pairwise Pearson correlations of species biomass across sampling units identified redundant species with similar spatial patterns. Species pairs with correlation coefficients >0.7 were flagged as potentially redundant, as highly correlated species provide minimal independent information but increase model complexity.


```{r, message=FALSE, warning=FALSE}
# Prepare wide format for correlation analysis
biomass_wide <- data_2025 %>%
  select(PRIMARY_SAMPLE_UNIT, SPECIES_CD, biomass) %>%
  pivot_wider(names_from = SPECIES_CD, 
              values_from = biomass, 
              values_fill = 0)

# Calculate correlation matrix
species_names <- setdiff(names(biomass_wide), "PRIMARY_SAMPLE_UNIT")
cor_matrix <- cor(biomass_wide %>% select(-PRIMARY_SAMPLE_UNIT), 
                  use = "pairwise.complete.obs")

cat("\n=== CORRELATION MATRIX SUMMARY ===\n")
cat("Number of species:", nrow(cor_matrix), "\n")
cat("Number of pairwise comparisons:", (nrow(cor_matrix)^2 - nrow(cor_matrix))/2, "\n")

# Check for NA/NaN values
n_na <- sum(is.na(cor_matrix))
if(n_na > 0) {
  cat("WARNING:", n_na, "NA values in correlation matrix\n")
  cat("These occur when species have zero variance (constant biomass)\n\n")
}

# Identify highly correlated pairs (r > 0.7)
high_cor_indices <- which(abs(cor_matrix) > 0.7 & cor_matrix != 1, arr.ind = TRUE)

if(nrow(high_cor_indices) > 0) {
  # Remove duplicate pairs (correlation matrix is symmetric)
  high_cor_indices <- high_cor_indices[high_cor_indices[,1] < high_cor_indices[,2], , drop = FALSE]
  
  high_cor_pairs <- data.frame(
    species1 = rownames(cor_matrix)[high_cor_indices[,1]],
    species2 = colnames(cor_matrix)[high_cor_indices[,2]],
    correlation = cor_matrix[high_cor_indices]
  ) %>%
    arrange(desc(abs(correlation)))
  
  cat("=== HIGHLY CORRELATED SPECIES PAIRS (|r| > 0.7) ===\n")
  print(high_cor_pairs, row.names = FALSE)
  cat("\nNumber of redundant pairs:", nrow(high_cor_pairs), "\n")
} else {
  cat("No species pairs with |r| > 0.7 detected\n")
  cat("This suggests species have distinct spatial patterns\n")
}

# Clean correlation matrix for plotting (replace NA with 0)
cor_matrix_clean <- cor_matrix
cor_matrix_clean[is.na(cor_matrix_clean)] <- 0

# Visualize correlation matrix with corrplot
corrplot(cor_matrix_clean, 
         method = "circle", 
         type = "upper",
         order = "original",  # Changed from "hclust" to avoid error
         tl.cex = 0.6, 
         tl.col = "black",
         tl.srt = 45,
         col = colorRampPalette(c("#6D9EC1", "white", "#E46726"))(200),
         title = "Species Biomass Correlation Matrix (2025)",
         mar = c(0,0,2,0),
         na.label = "square",
         na.label.col = "gray90",
         diag = FALSE)

# Alternative: Only plot species with valid correlations
valid_species <- apply(cor_matrix, 1, function(x) sum(is.na(x)) == 0)
if(sum(valid_species) > 1) {
  cor_matrix_valid <- cor_matrix[valid_species, valid_species]
  
  # Try hierarchical clustering only if no NAs
  if(sum(is.na(cor_matrix_valid)) == 0) {
    library(pheatmap)
    
    pheatmap(cor_matrix_valid,
             clustering_distance_rows = "euclidean",
             clustering_distance_cols = "euclidean",
             clustering_method = "complete",
             color = colorRampPalette(c("#6D9EC1", "white", "#E46726"))(100),
             border_color = NA,
             fontsize = 8,
             main = "Hierarchical Clustering of Species by Biomass Correlation (2025)",
             display_numbers = FALSE,
             angle_col = 45)
  } else {
    cat("\nSkipping hierarchical clustering due to NA values in correlation matrix\n")
  }
}

# Distribution of correlation coefficients (excluding NA)
cor_vector <- cor_matrix[upper.tri(cor_matrix)]
cor_vector <- cor_vector[!is.na(cor_vector)]

if(length(cor_vector) > 0) {
  cat("\n=== CORRELATION DISTRIBUTION ===\n")
  summary(cor_vector)
  
  hist(cor_vector, 
       breaks = 30, 
       main = "Distribution of Pairwise Species Correlations (2025)",
       xlab = "Pearson Correlation Coefficient",
       ylab = "Frequency",
       col = "steelblue",
       border = "white")
  abline(v = 0.7, col = "red", lwd = 2, lty = 2)
  abline(v = -0.7, col = "red", lwd = 2, lty = 2)
  abline(v = 0, col = "gray30", lwd = 1, lty = 1)
  legend("topright", 
         legend = c("r = ±0.7 threshold", "No correlation"),
         col = c("red", "gray30"),
         lty = c(2, 1),
         lwd = c(2, 1))
}

# Identify species with most correlations > 0.5 (moderately correlated)
if(sum(!is.na(cor_matrix)) > 0) {
  cor_counts <- apply(cor_matrix, 1, function(x) {
    sum(abs(x) > 0.5 & abs(x) < 1, na.rm = TRUE)
  })
  
  cat("\n=== SPECIES WITH MOST MODERATE CORRELATIONS (|r| > 0.5) ===\n")
  sort(cor_counts, decreasing = TRUE)[1:10] %>% print()
}
```
Correlation analysis revealed no species pairs with correlation coefficients exceeding |r| = 0.7, indicating that species exhibit largely independent spatial patterns across the study area. A total of 231 pairwise comparisons (29.6%) resulted in undefined correlations due to species having zero variance (constant or absent across all sites), further confirming the presence of hyper-rare, uninformative species in the dataset. Only two species (*Stegastes partitus* and *Thalassoma bifasciatum*) showed moderate correlation (|r| > 0.5) with other species, both of which were retained in the final selection due to their high quality metrics. The lack of strong inter-species correlations confirmed that removing low-quality species based on variability criteria would not result in loss of independent spatial information.



## 5. Zero-Inflation Analysis

The proportion of zero observations (absence records) was calculated for each species. High zero-inflation (>80%) indicates rare or highly patchy distributions that inflate variance and complicate statistical modeling. Zero-inflated species often require specialized models and may not aggregate well with continuously present species.

```{r, message=FALSE, warning=FALSE}
# Calculate proportion of zeros
zero_inflation <- data_2025 %>%
  group_by(group, SPECIES_CD, SPECIES) %>%
  summarise(
    n_obs = n(),
    n_zeros = sum(biomass == 0),
    prop_zeros = (n_zeros / n_obs) * 100,
    mean_when_present = mean(biomass[biomass > 0], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(group, prop_zeros)

# Summary statistics
cat("\n=== ZERO-INFLATION SUMMARY ===\n")
summary(zero_inflation$prop_zeros)

# Plot: Zero-inflation vs biomass when present
ggplot(zero_inflation, aes(x = prop_zeros, y = mean_when_present, 
                           color = group, label = SPECIES_CD)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_vline(xintercept = 80, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 75, y = max(zero_inflation$mean_when_present, na.rm = TRUE), 
           label = "80% zeros threshold", color = "red", angle = 90, vjust = -0.5) +
  geom_text_repel(size = 2.5, max.overlaps = 20) +
  scale_y_log10() +
  theme_classic() +
  labs(title = "Zero-Inflation vs Mean Biomass When Present (2025)",
       subtitle = "Species in upper-left are rare but important when present",
       x = "% Zeros (absence)", 
       y = "Mean Biomass When Present (kg, log10)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10),
        legend.position = "bottom")
```

## 6. Taylor's Power Law

We examined the mean-variance relationship following Taylor's Power Law: log(variance) = log(a) + b × log(mean). The slope parameter (b) indicates spatial aggregation patterns, where b ≈ 2 suggests random distribution, b > 2 indicates patchy/aggregated distribution, and b < 2 suggests regular spacing. Species deviating strongly from this relationship represent outliers in their distributional ecology.

```{r, message=FALSE, warning=FALSE}
# Combine mean and variance data
taylor_data <- species_cv %>%
  select(group, SPECIES_CD, SPECIES, mean_biomass, sd_biomass) %>%
  mutate(
    variance = sd_biomass^2,
    log_mean = log10(mean_biomass + 0.001),  # Add small constant to avoid log(0)
    log_var = log10(variance + 0.001)
  ) %>%
  filter(is.finite(log_mean) & is.finite(log_var))

# Fit Taylor's Power Law: log(variance) = log(a) + b * log(mean)
taylor_model <- lm(log_var ~ log_mean, data = taylor_data)

cat("\n=== TAYLOR'S POWER LAW ===\n")
cat("Model: log(variance) = a + b * log(mean)\n\n")
summary(taylor_model)

# Extract slope (b)
slope <- coef(taylor_model)[2]
intercept <- coef(taylor_model)[1]

cat("\nInterpretation of slope (b =", round(slope, 2), "):\n")
if(slope < 1.8) {
  cat("  b < 2: Species tend toward regular spacing\n")
} else if(slope > 2.2) {
  cat("  b > 2: Species are aggregated/patchy\n")
} else {
  cat("  b ≈ 2: Species show random spatial distribution\n")
}

# Visualize Taylor's Power Law
ggplot(taylor_data, aes(x = log_mean, y = log_var, color = group, label = SPECIES_CD)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  geom_text_repel(size = 2.5, max.overlaps = 15) +
  annotate("text", 
           x = min(taylor_data$log_mean, na.rm = TRUE) + 0.5,
           y = max(taylor_data$log_var, na.rm = TRUE) - 0.5,
           label = paste0("slope = ", round(slope, 2), 
                         "\nR² = ", round(summary(taylor_model)$r.squared, 3)),
           hjust = 0, size = 4, fontface = "bold") +
  theme_classic() +
  labs(title = "Taylor's Power Law: Mean-Variance Relationship (2025)",
       subtitle = "Species above the line are more variable than expected",
       x = "log10(Mean Biomass)",
       y = "log10(Variance)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10),
        legend.position = "bottom")

# Identify outliers (residuals > 2 SD)
taylor_data <- taylor_data %>%
  mutate(
    fitted = predict(taylor_model),
    residuals = log_var - fitted,
    outlier = abs(residuals) > 2 * sd(residuals)
  )

cat("\n=== SPECIES DEVIATING FROM MEAN-VARIANCE RELATIONSHIP ===\n")
cat("(Residuals > 2 SD - higher variance than expected)\n\n")
taylor_data %>%
  filter(outlier) %>%
  select(group, SPECIES, mean_biomass, variance, residuals) %>%
  arrange(desc(residuals)) %>%
  print()
```

# Decision Framework ---------------------------------------------------------

## Integrate All Metrics

```{r}
# Combine all metrics into single dataframe
species_metrics <- species_cv %>%
  left_join(species_frequency %>% select(SPECIES_CD, freq_occurrence), 
            by = "SPECIES_CD") %>%
  left_join(species_contribution %>% select(SPECIES_CD, mean_contribution), 
            by = "SPECIES_CD") %>%
  left_join(zero_inflation %>% select(SPECIES_CD, prop_zeros), 
            by = "SPECIES_CD")

# Calculate composite selection score
species_metrics <- species_metrics %>%
  mutate(
    selection_score = 
      (100 - pmin(cv, 500)) * 0.30 +      # Cap CV at 500 for scoring
      freq_occurrence * 0.25 +
      mean_contribution * 0.25 +
      (100 - prop_zeros) * 0.20
  ) %>%
  arrange(group, desc(selection_score))

cat("\n=== INTEGRATED SPECIES METRICS SUMMARY ===\n")
summary(species_metrics %>% select(cv, freq_occurrence, mean_contribution, prop_zeros, selection_score))

# Display top 20 species by selection score
cat("\n=== TOP 20 SPECIES BY SELECTION SCORE ===\n")
species_metrics %>% 
  arrange(desc(selection_score)) %>% 
  head(20) %>%
  select(group, SPECIES, cv, freq_occurrence, mean_contribution, prop_zeros, selection_score) %>%
  print(n = 20)
```

## Three-Tier Selection Strategy

```{r}
# Tier 1: High quality species
tier1 <- species_metrics %>%
  filter(
    !is.na(cv),
    cv < 350,
    freq_occurrence > 15,
    mean_contribution > 10,
    prop_zeros < 80
  ) %>%
  mutate(selection_method = "High quality")

cat("\n=== TIER 1: HIGH QUALITY SPECIES ===\n")
cat("Criteria: CV < 350, Frequency > 15%, Contribution > 10%, Zeros < 80%\n")
cat("Species selected:", nrow(tier1), "\n\n")
print(tier1 %>% select(group, SPECIES, cv, freq_occurrence, mean_contribution))

# Tier 2: Dominant species (>30% of family biomass)
tier2 <- species_metrics %>%
  filter(!is.na(cv)) %>%
  anti_join(tier1, by = "SPECIES_CD") %>%
  filter(mean_contribution > 30) %>%
  mutate(selection_method = "Dominant")

cat("\n=== TIER 2: DOMINANT SPECIES ===\n")
cat("Criteria: Contribution > 30% (regardless of other metrics)\n")
cat("Species selected:", nrow(tier2), "\n\n")
print(tier2 %>% select(group, SPECIES, cv, freq_occurrence, mean_contribution))

# Tier 3: Best available per family
tier3 <- species_metrics %>%
  filter(!is.na(cv)) %>%
  anti_join(bind_rows(tier1, tier2), by = "SPECIES_CD") %>%
  group_by(group) %>%
  slice_max(order_by = selection_score, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(selection_method = "Best available")

cat("\n=== TIER 3: BEST AVAILABLE (family representatives) ===\n")
cat("Criteria: Highest selection score among remaining species per family\n")
cat("Species selected:", nrow(tier3), "\n\n")
print(tier3 %>% select(group, SPECIES, cv, freq_occurrence, mean_contribution))

# Combine all tiers
preliminary_selection <- bind_rows(tier1, tier2, tier3) %>%
  arrange(group, desc(selection_score))

cat("\n=== PRELIMINARY SELECTION SUMMARY ===\n")
cat("Total species:", nrow(preliminary_selection), "\n")
cat("Families represented:", n_distinct(preliminary_selection$group), "/", 
    n_distinct(species_metrics$group), "\n\n")

preliminary_selection %>%
  group_by(selection_method) %>%
  summarise(
    n_species = n(),
    mean_cv = mean(cv, na.rm = TRUE),
    mean_freq = mean(freq_occurrence),
    mean_contrib = mean(mean_contribution),
    mean_zeros = mean(prop_zeros)
  ) %>%
  print()
```

### Remove Hyper-Rare Apogonids

```{r, message=FALSE, warning=FALSE}
# Identify problematic Apogonids
problematic_apogonids <- preliminary_selection %>%
  filter(
    group == "Apogonidae",
    selection_method == "Dominant",
    cv > 2000 | freq_occurrence < 2
  )

cat("\n=== PROBLEMATIC SPECIES IDENTIFIED FOR REMOVAL ===\n")
cat("Criteria: Apogonidae with CV > 2000% or Frequency < 2%\n\n")
print(problematic_apogonids %>% 
        select(SPECIES, cv, freq_occurrence, mean_contribution, prop_zeros))

# Final selection: Remove hyper-rare Apogonids
final_selection <- preliminary_selection %>%
  filter(
    !(SPECIES %in% c("Apogon binotatus", "Apogon quadrisquamatus"))
  )

cat("\n\n=== FINAL SPECIES SELECTION (OPTION 2: MODERATE) ===\n")
cat("Total species selected:", nrow(final_selection), "\n")
cat("Families represented:", n_distinct(final_selection$group), "\n")
cat("Species removed: 2 hyper-rare Apogonids\n\n")

# Final summary by tier
final_selection %>%
  group_by(selection_method) %>%
  summarise(
    n_species = n(),
    mean_cv = round(mean(cv, na.rm = TRUE), 1),
    mean_freq = round(mean(freq_occurrence), 1),
    mean_contrib = round(mean(mean_contribution), 1),
    mean_zeros = round(mean(prop_zeros), 1)
  ) %>%
  print()

# Complete final list
cat("\n=== COMPLETE FINAL SPECIES LIST ===\n")
final_selection %>%
  select(group, SPECIES, selection_method, cv, freq_occurrence, 
         mean_contribution, prop_zeros, selection_score) %>%
  arrange(selection_method, group, desc(selection_score)) %>%
  print(n = Inf)

# Export final selection
write.csv(final_selection, 
          "final_selected_species_2025.csv", 
          row.names = FALSE)

cat("\nFinal species list exported to: final_selected_species_2025.csv\n")
```


## Variance Reduction Validation

```{r}
cat("\n=== VARIANCE REDUCTION VALIDATION ===\n\n")

# Get selected species codes
selected_species_codes <- final_selection$SPECIES_CD

# 1. Calculate family-level biomass with ALL species
all_species_biomass <- data_2025 %>%
  group_by(PRIMARY_SAMPLE_UNIT, group, REGION, STRAT, PROT, 
           LAT_DEGREES, LON_DEGREES, DEPTH) %>%
  summarise(
    family_biomass = sum(biomass, na.rm = TRUE),
    n_species_present = n_distinct(SPECIES_CD[biomass > 0]),
    .groups = "drop"
  ) %>%
  mutate(dataset = "All species (n=40)")

# 2. Calculate family-level biomass with SELECTED species only
selected_species_biomass <- data_2025 %>%
  filter(SPECIES_CD %in% selected_species_codes) %>%
  group_by(PRIMARY_SAMPLE_UNIT, group, REGION, STRAT, PROT, 
           LAT_DEGREES, LON_DEGREES, DEPTH) %>%
  summarise(
    family_biomass = sum(biomass, na.rm = TRUE),
    n_species_present = n_distinct(SPECIES_CD[biomass > 0]),
    .groups = "drop"
  ) %>%
  mutate(dataset = "Selected species (n=24)")

# Combine datasets
comparison_data <- bind_rows(all_species_biomass, selected_species_biomass)

# Calculate summary statistics by family and dataset
variance_comparison <- comparison_data %>%
  group_by(group, dataset) %>%
  summarise(
    n_sites = n(),
    mean_biomass = mean(family_biomass, na.rm = TRUE),
    sd_biomass = sd(family_biomass, na.rm = TRUE),
    cv = (sd_biomass / mean_biomass) * 100,
    median_biomass = median(family_biomass, na.rm = TRUE),
    prop_zeros = sum(family_biomass == 0) / n() * 100,
    .groups = "drop"
  )

# Calculate variance reduction
variance_reduction <- variance_comparison %>%
  select(group, dataset, cv, sd_biomass) %>%
  pivot_wider(names_from = dataset, 
              values_from = c(cv, sd_biomass)) %>%
  mutate(
    cv_reduction_pct = ((`cv_All species (n=40)` - `cv_Selected species (n=24)`) / 
                          `cv_All species (n=40)`) * 100,
    sd_reduction_pct = ((`sd_biomass_All species (n=40)` - `sd_biomass_Selected species (n=24)`) / 
                          `sd_biomass_All species (n=40)`) * 100
  )

cat("=== VARIANCE REDUCTION BY FAMILY ===\n\n")
variance_reduction %>%
  select(group, `cv_All species (n=40)`, `cv_Selected species (n=24)`, 
         cv_reduction_pct) %>%
  arrange(desc(cv_reduction_pct)) %>%
  print(n = Inf)

cat("\n=== OVERALL VARIANCE REDUCTION ===\n")
overall_reduction <- variance_comparison %>%
  group_by(dataset) %>%
  summarise(
    mean_cv = mean(cv, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    mean_sd = mean(sd_biomass, na.rm = TRUE)
  )
print(overall_reduction)

overall_improvement <- overall_reduction %>%
  summarise(
    cv_reduction = (mean_cv[1] - mean_cv[2]) / mean_cv[1] * 100,
    sd_reduction = (mean_sd[1] - mean_sd[2]) / mean_sd[1] * 100
  )

cat("\nMean CV reduction:", round(overall_improvement$cv_reduction, 1), "%\n")
cat("Mean SD reduction:", round(overall_improvement$sd_reduction, 1), "%\n")

# Visualize variance comparison
ggplot(variance_comparison, aes(x = reorder(group, -cv), y = cv, fill = dataset)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme_classic() +
  labs(title = "Coefficient of Variation: All vs Selected Species",
       subtitle = "Lower CV indicates reduced variability in family biomass estimates",
       x = "Family",
       y = "CV (%)",
       fill = "Dataset") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10),
        legend.position = "bottom") +
  scale_fill_manual(values = c("All species (n=40)" = "#E69F00", 
                                "Selected species (n=24)" = "#56B4E9"))

# Biomass comparison scatter plot
biomass_comparison_wide <- comparison_data %>%
  select(PRIMARY_SAMPLE_UNIT, group, dataset, family_biomass) %>%
  pivot_wider(names_from = dataset, values_from = family_biomass)

ggplot(biomass_comparison_wide, 
       aes(x = `All species (n=40)`, y = `Selected species (n=24)`)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  facet_wrap(~ group, scales = "free") +
  theme_classic() +
  labs(title = "Family Biomass: All Species vs Selected Species",
       subtitle = "Points near the 1:1 line indicate biomass is preserved",
       x = "Biomass - All Species (kg)",
       y = "Biomass - Selected Species (kg)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 10))

# Calculate biomass retention (how much biomass is preserved)
biomass_retention <- comparison_data %>%
  select(PRIMARY_SAMPLE_UNIT, group, dataset, family_biomass) %>%
  pivot_wider(names_from = dataset, values_from = family_biomass) %>%
  group_by(group) %>%
  summarise(
    total_biomass_all = sum(`All species (n=40)`, na.rm = TRUE),
    total_biomass_selected = sum(`Selected species (n=24)`, na.rm = TRUE),
    biomass_retained_pct = (total_biomass_selected / total_biomass_all) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(biomass_retained_pct))

cat("\n=== BIOMASS RETENTION BY FAMILY ===\n")
cat("(% of total biomass retained after species selection)\n\n")
print(biomass_retention, n = Inf)

cat("\nOverall biomass retention:", 
    round(sum(biomass_retention$total_biomass_selected) / 
            sum(biomass_retention$total_biomass_all) * 100, 1), "%\n")

# Summary table for publication
summary_table <- variance_reduction %>%
  left_join(biomass_retention, by = "group") %>%
  select(group, 
         cv_all = `cv_All species (n=40)`, 
         cv_selected = `cv_Selected species (n=24)`,
         cv_reduction_pct,
         biomass_retained_pct) %>%
  arrange(desc(cv_reduction_pct))

cat("\n=== FINAL SUMMARY TABLE ===\n")
print(summary_table, n = Inf)

# Export summary
write.csv(summary_table, 
          "variance_reduction_summary_2025.csv", 
          row.names = FALSE)

cat("\nVariance reduction summary exported to: variance_reduction_summary_2025.csv\n")
```

## Species Selection and Variance Reduction -----------------------------------

**Taylor's Power Law and Spatial Distribution Patterns**

Analysis of the mean-variance relationship across all 40 species revealed a significant adherence to Taylor's Power Law (R² = 0.83, F₁,₃₈ = 184.6, p < 0.001). The estimated slope (b = 0.92, SE = 0.067) was significantly less than 2, indicating that reef fish species in this assemblage tend toward regular spacing rather than aggregated distributions. This relatively low slope suggests lower spatial variance than expected under random distribution, potentially reflecting territorial behaviors or competitive interactions that promote spatial regularity. One species, Lutjanus apodus (schoolmaster snapper), exhibited significantly higher variance than predicted by the mean-variance relationship (residual > 2 SD), indicating unusually high spatial patchiness relative to its mean biomass.

**Integrated Species Metrics**

The 40 candidate species exhibited substantial variation in all metrics (Table X). Coefficient of variation ranged from 111% to 2,579% (median = 322%), reflecting the naturally patchy distribution of reef fishes. Frequency of occurrence ranged from 0% to 95% (median = 30%), with many species present at fewer than one-third of sampling sites. Species contributions to family biomass ranged from effectively zero to 100% (median = 14.6%), indicating that most families were dominated by a few key species. Zero-inflation was high across the assemblage (median = 70% zeros), confirming that most species were absent from most sites. The composite selection score, integrating all metrics, ranged from -120 to +47, with only the upper quartile of species achieving positive scores.

Correlation analysis revealed independent spatial patterns among species, with no species pairs exhibiting correlations exceeding |r| = 0.7. However, 29.6% of pairwise correlations were undefined due to species having zero variance (constant absence or presence at identical biomass levels), confirming the prevalence of hyper-rare, uninformative species in the initial dataset.

**Three-Tier Species Selection**

Tier 1 (High Quality Species, n = 16): Sixteen species met all stringent criteria (CV < 350%, frequency > 15%, contribution > 10%, zeros < 80%). These species exhibited exceptional stability (mean CV = 201%), high detection rates (mean frequency = 62%), and substantial ecological importance (mean contribution = 51% of family biomass). This tier included ecologically dominant Caribbean reef species such as the parrotfishes Sparisoma aurofrenatum (redband parrotfish, CV = 111%) and Sparisoma viride (stoplight parrotfish, CV = 156%), the wrasses Thalassoma bifasciatum (bluehead wrasse, CV = 137%) and Halichoeres garnoti (yellowhead wrasse, CV = 143%), and the damselfish Stegastes partitus (bicolor damselfish, CV = 195%). These species represented the modeling core, providing stable, predictable biomass estimates across the study area.

Tier 2 (Dominant Species, n = 1): One species, Lutjanus apodus (schoolmaster snapper), contributed >30% of family biomass (74%) despite moderate variability (CV = 394%) and intermediate occurrence (26% of sites). This species was retained due to its overwhelming dominance of Lutjanidae biomass, ensuring the family's ecological importance was adequately represented despite spatial patchiness.

Tier 3 (Best Available Representatives, n = 7): Seven families lacking Tier 1 or 2 representatives were represented by their single highest-scoring species. These included Apogon maculatus (flamefish, Apogonidae), Chaetodon ocellatus (spotfin butterflyfish, Chaetodontidae), Coryphopterus glaucofraenum (bridled goby, Gobiidae), Myripristis jacobus (blackbar soldierfish, Holocentridae), Halichoeres maculipinna (clown wrasse, Labridae), Lutjanus mahogoni (mahogany snapper, Lutjanidae), and Stegastes adustus (dusky damselfish, Pomacentridae). Although these species exhibited higher variability (mean CV = 598%) and lower occurrence (mean frequency = 20%), they ensured complete taxonomic coverage across all 11 families.

Exclusion of Hyper-Rare Apogonids: Two extremely rare Apogonidae species (Apogon binotatus and A. quadrisquamatus) initially selected as Tier 2 dominants were excluded from the final list due to prohibitive instability (CV > 2,000%) and near-complete absence (occurrence < 2%, >98% zeros). Their removal reduced Apogonidae representation to a single species but eliminated species contributing more variance than information.

**Final Selection and Variance Reduction**
The final selection comprised 24 species across 11 families, down from the initial 40 species. Species selection achieved an overall 7.5% reduction in mean coefficient of variation (from 317% to 294%) and a 2.8% reduction in mean standard deviation while retaining 96.2% of total biomass (Table X).

Variance reduction was highly variable among families. The most substantial improvement occurred in Apogonidae (21% CV reduction) following removal of the two hyper-rare species. Seven families (Epinephelidae, Gobiidae, Haemulidae, Holocentridae, Scaridae, Tetraodontidae) exhibited no change in CV because all species were retained (single-species families or all species meeting criteria). Conversely, four families showed slight CV increases: Chaetodontidae (+0.4%), Labridae (+2.7%), Pomacentridae (+7.6%), and Lutjanidae (+7.8%). These increases occurred because removed species, despite high variability, occasionally contributed biomass that dampened family-level variance when aggregated. However, all four families retained >86% of total biomass, and the trade-off between modest CV increases and improved model parsimony (removing 16 species) was deemed acceptable.

The modest overall variance reduction (7.5%) reflects the biological reality that retained species themselves exhibit high natural variability (median CV = 293%). This indicates that reef fish patchiness is not primarily driven by rare, erratic species but is an inherent property of the assemblage. The species selection process successfully eliminated uninformative species while preserving ecological information, as evidenced by near-complete biomass retention (96.2%).

**Interpretation**
The final 24-species selection balances ecological representativeness, statistical stability, and model parsimony. High-quality species (n = 16) provide the stable foundation for spatial prediction, while dominant and representative species (n = 8) ensure no family is ecologically misrepresented. The lack of strong inter-species correlations (maximum |r| < 0.7) confirms that each retained species contributes independent spatial information to family-level biomass estimates.

The limited variance reduction achieved (7.5%) should not be interpreted as selection failure. Rather, it demonstrates that the majority of spatial variability is intrinsic to reef fish ecology, not an artifact of including rare species. Caribbean reef fish assemblages are naturally patchy due to habitat heterogeneity, behavioral territoriality (evidenced by Taylor's slope < 2), and recruitment variability. The species selection process successfully identified this intrinsic variability and removed only those species contributing disproportionate noise (231 undefined correlations, >98% zeros) while preserving the ecological signal (96% biomass retention).

## Table
```{r, results='asis'}
# Load packages
library(knitr)
library(kableExtra)
library(dplyr)
library(tibble)

# First, verify the input data exists
cat("Checking data objects...\n")
cat("variance_reduction rows:", nrow(variance_reduction), "\n")
cat("biomass_retention rows:", nrow(biomass_retention), "\n")

# Prepare publication table data
pub_table <- variance_reduction %>%
  left_join(biomass_retention %>% select(group, biomass_retained_pct), 
            by = "group") %>%
  left_join(
    species_metrics %>% 
      count(group, name = "n_total"),
    by = "group"
  ) %>%
  left_join(
    final_selection %>% 
      count(group, name = "n_selected"),
    by = "group"
  ) %>%
  mutate(
    species_counts = paste0(n_total, "/", n_selected),
    cv_all = round(`cv_All species (n=40)`, 1),
    cv_selected = round(`cv_Selected species (n=24)`, 1),
    cv_reduction = round(cv_reduction_pct, 1),
    biomass_retained = round(biomass_retained_pct, 1)
  ) %>%
  select(
    Family = group,
    `Species (all/selected)` = species_counts,
    `CV All (%)` = cv_all,
    `CV Selected (%)` = cv_selected,
    `CV Reduction (%)` = cv_reduction,
    `Biomass Retained (%)` = biomass_retained
  ) %>%
  arrange(Family)

# Check if pub_table has data
cat("pub_table rows:", nrow(pub_table), "\n")

# Add overall row
overall_row <- data.frame(
  Family = "Overall",
  `Species (all/selected)` = "40/24",
  `CV All (%)` = round(mean(variance_reduction$`cv_All species (n=40)`, na.rm = TRUE), 1),
  `CV Selected (%)` = round(mean(variance_reduction$`cv_Selected species (n=24)`, na.rm = TRUE), 1),
  `CV Reduction (%)` = round(
    (mean(variance_reduction$`cv_All species (n=40)`, na.rm = TRUE) - 
     mean(variance_reduction$`cv_Selected species (n=24)`, na.rm = TRUE)) / 
    mean(variance_reduction$`cv_All species (n=40)`, na.rm = TRUE) * 100, 1),
  `Biomass Retained (%)` = round(
    sum(biomass_retention$total_biomass_selected, na.rm = TRUE) / 
    sum(biomass_retention$total_biomass_all, na.rm = TRUE) * 100, 1),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

pub_table_final <- bind_rows(pub_table, overall_row)

# Print dimensions for debugging
cat("Final table dimensions:", nrow(pub_table_final), "x", ncol(pub_table_final), "\n\n")

# Create the table
kbl(pub_table_final,
    format = "latex",
    booktabs = TRUE,
    caption = "Summary of species selection criteria and variance reduction achieved by taxonomic family. CV = Coefficient of Variation.",
    align = c("l", "c", "c", "c", "c", "c"),
    linesep = "",
    longtable = FALSE) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size = 10,
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(nrow(pub_table_final), bold = TRUE, hline_after = TRUE) %>%
  footnote(
    general = "Families showing CV increases retained >86\\\\% of biomass, indicating trade-off between model parsimony and minimal information loss.",
    general_title = "Note:",
    footnote_as_chunk = TRUE
  )
```



